Detection and replication of epistasis influencing transcription in humans

### 2013-06-18 docs/manuscript/nature_letter.tex
manuscript updated

Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power. For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects. When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes. Because our focus was to ascertain the extent to which epistasis exists amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes. These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that many genetic effects will be relatively large, maximising the chance at detecting epistasis, should it exist.

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood. After stringent filtering and multiple testing correction (Methods) we had identified 501 putative genetic interactions influencing 238 gene expression levels. Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}). These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).

Additionally, there was extreme enrichment for interaction effects among the discovery SNPs that did not reach the stringent significance threshold for replication (Figure \ref{fig:qqMeta}). We observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values exceeding the one-tailed 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Supplementary Figure \ref{fig:qqMetaNonsig}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent populations. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication. It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is theoretically difficult because the dependence on LD between observed SNPs and causal variants is on average three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}. Therefore these results are very encouraging with regards to the detection and replication of epistasis.

@@

We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP. For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect. We observed that nine of the 14 masking SNPs are located in intronic regions (proportion of SNP panel in introns $= 0.05$, $p = 3.11 \times 10^{-9}$). Each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).

% \subsection{Functional annotation}
Of the discovery interactions, 47 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 441 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting. In total the 501 interactions comprised 781 unique SNPs, which we analysed for functional enrichment. We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013}. There was significant enrichment for \emph{cis}-acting SNPs in haematopoeitic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing, Supplementary Figure \ref{fig:transh3k4me3}). However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues, Supplementary Figure \ref{fig:transh3k4me3}). This difference between \emph{cis} and \emph{trans} SNPs suggests that there is a range of molecular mechanisms by which epistasis might arise and the \emph{cis}-SNPs may provide tissue specificity in these interactions. There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a} (Supplementary Figure \ref{fig:enhancers}). This enrichment is consistent for both \emph{cis} and \emph{trans} SNPs ($p < 1 \times 10^{-6}$). In particular, there was substantial enrichment for the GATA2 binding motif within 1kb of all epistatic SNPs, a known regulator of transcription in haematopoietic cells \cite{Tsai1994} ($p = 1 \times 10^{-40}$).

We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}. We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}). Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.

% \subsection{Contribution of epistatic variance}
Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question. In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold. How does this compare to the number of traits controlled by additive effects? The BSGS dataset has been previously analysed for additive effects at all expression traits \cite{Powell2012}, and if we take all the additive eQTLs that were significant at the epistatic threshold of $p < 2.91 \times 10^{-16}$ we find that 453 gene expression levels out of the 7339 analysed had at least one significant expression quantitative trait locus (eQTL). Therefore it can be argued that the number of instances of detectable epistasis are substantial.

However in terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Ideally one would approach this question from a whole genome level \cite{Visscher2008} but this is intractable for non-additive variance components. Yet some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}. Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that in total they explain 1.73\% of the total phenotypic variance of all 7339 probes. By contrast, the epistatic variance from the interacting SNPs detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance. This is broadly in line with theory \cite{Hill2008a}, but there are three caveats to this comparison. Firstly, the ratio of additive to epistatic variance may differ at different effect sizes. Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{Grey_graph_of_interactions_2_lists.pdf}
	\caption{\textbf{Replication of epistatic networks in two independent cohorts} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown. 316 Interactions that had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data are shown with black nodes for SNPs and red nodes for gene expression traits. The interactions that fell within the confidence interval and showed no evidence of replication are greyed out. There is good congruence of interactions between independent populations.}
	\label{fig:fireworks}
\end{figure}
\clearpage

\begin{figure}[H]
	\centering
	\includegraphics[width=2.5in]{gpbonfrep.pdf}
	\caption{\textbf{Replication of genotype-phenotype (GP) maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown. Each GP map consists of nine tiles where each tile represents the expression level for that two-locus genotype class. Phenotypes are for gene transcript levels (dark coloured tiles = low expression, light coloured tiles = high expression). Columns of GP maps are for each independent population. Rows of GP maps are for each of 30 significantly replicated interactions, corresponding to the rows in Table \ref{tab:bonferroni}. There is a general trend of the GP maps replicating across all three datasets.}
	\label{fig:gpmaps}
\end{figure}
\clearpage

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{cis_h3k4me3}
	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3. Here we show that there is }
	\label{fig:cish3k4me3}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=5in]{trans_h3k4me3}
	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3. Here we show that there is }
	\label{fig:transh3k4me3}
\end{figure}
\clearpage

>>>>

Methods are now available to overcome the computational problems involved in searching for epistasis, but its detection still remains problematic due to reduced statistical power. For example increased dependence on linkage disequilibrium (LD) between causal SNPs and observed SNPs \cite{Weir2008, Hemani2013}, increased model complexity in fitting interaction terms \cite{Marchini2005}, and more extreme significance thresholds to account for increased multiple testing \cite{Cordell2009} all make it more difficult to detect epistasis in comparison to additive effects. When genetic effect sizes are small, as is expected in most complex traits of interest \cite{Visscher2012}, the power to detect epistasis diminishes rapidly. There are two simple ways to overcome this problem. One is by using extremely large sample sizes \cite{LangoAllen2010}; another is by analysing traits that are likely to have large effect sizes. Because our focus was to ascertain the extent to which instances of epistasis occur amongst natural genetic variation we designed a study around the latter approach and searched for epistatic genetic effects that influence gene expression levels. Transcription levels can be measured for thousands of genes. These traits are largely heritable but on average less polygenic than high level phenotypes \cite{Powell2013}, thus it is expected that many genetic effects will be relatively large, maximising the chance at detecting epistasis, should it exist.

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood. After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing 238 gene expression levels. Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}). These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).

Additionally, there was enrichment for interaction effects among the discovery SNPs that did not reach the stringent significance threshold for replication (Figure \ref{fig:qqMeta}). We observed that 316 of the remaining 404 discovery SNPs had replication interaction $p$-values exceeding the one-tailed 2.5\% confidence interval under the null distribution of no effects ($p << 1.0 \times 10^{-16}$, Supplementary Figure \ref{fig:qqMetaNonsig}). The congruence of the epistatic networks in discovery and replication datasets is shown in Figure \ref{fig:fireworks}, demonstrating that these complex genetic patterns are common even across independent populations. A further replication was attempted using the Centre for Health Discovery and Wellbeing (CHDWB) dataset \cite{Preininger2013}, but only 185 of the SNP pairs passed filtering because the sample size was small ($n=139$), and due to insufficient power we found no evidence for replication. It should be noted that although it is a necessary step to establish the veracity of the signals from the discovery set, replication of epistasis is theoretically difficult because the dependence on LD between observed SNPs and causal variants is on average three orders of magnitude higher than it is for independent additive effects \cite{Weir2008, Hemani2013}. Therefore these results are very encouraging with regards to the detection and replication of epistasis.

@@

Of the discovery interactions, 47 were \emph{cis}-\emph{cis} acting (both SNPs were on the same chromosome as the expression gene), 441 were \emph{cis}-\emph{trans}-acting, and 13 were \emph{trans}-\emph{trans}-acting. We observed a wide range of significant GP maps (Figure \ref{fig:gpmaps}) but the most common pattern of epistasis that we detected involved a \emph{trans}-SNP masking the effect of an additive \emph{cis}-SNP. For example, MBNL1 (involved in RNA modification and regulation of splicing \cite{Ho2004}) has a \emph{cis} effect at rs13069559 which in turn is controlled by 13 \emph{trans}-SNPs and one \emph{cis}-SNP that each exhibit a masking pattern, such that when the \emph{trans}-SNP is homozygous for the masking allele the decreasing allele of the \emph{cis}-SNP no longer has an affect. We observed that nine of the 14 masking SNPs are located in intronic regions (proportion of SNP panel in introns $= 0.05$, $p = 3.11 \times 10^{-9}$, Supplementary Figure \ref{fig:snplocations}). Each of these interactions have evidence for replication in at least one dataset and six are significant at the Bonferroni level (Supplementary Figure \ref{fig:circleplots}).

% \subsection{Functional annotation}
In total the 501 interactions comprised 781 unique SNPs, which we analysed for functional enrichment (Methods). We tested the SNPs for cell-type specific overlap with transcriptionally active chromatin regions, tagged by histone-3-lysine-4,3-methylation (H3K4me3) chromatin marks, in 34 cell types \cite{Trynka2013} (Supplementary Figure \ref{fig:h3k4me3}). There was significant enrichment for \emph{cis}-acting SNPs in haematopoeitic cell types only ($p < 1 \times 10^{-4}$ for the three tissues with the strongest enrichment after adjusting for multiple testing). However \emph{trans}-acting SNPs did not show any tissue specific enrichment ($p > 0.1$ for all tissues). This difference between \emph{cis} and \emph{trans} SNPs suggests a range of molecular mechanisms by which epistasis might arise and the \emph{cis}-SNPs may provide tissue specificity in these interactions. There is also strong enrichment for SNPs to be localised in enhancer regions \cite{Ward2012a} (Supplementary Figure \ref{fig:enhancers}). This enrichment is consistent for both \emph{cis} and \emph{trans} SNPs ($p < 1 \times 10^{-6}$). In particular, there was substantial enrichment for the GATA2 binding motif within 1kb of all epistatic SNPs, a known regulator of transcription in haematopoietic cells \cite{Tsai1994} ($p = 1 \times 10^{-40}$).

We also demonstrate a putative novel mechanism by which biological function can lead to epistatic genetic variance. It has been shown that different chromosomal regions spatially colocalise in the cell through chromatin interactions \cite{Lieberman-Aiden2009}. We cross referenced our epistatic SNPs with a map of chromosome interacting regions ($n = 96139$) in K562 blood cell lines \cite{Lan2012} (Methods) and found that 44 epistatic interactions mapped to within 2Mb ($p < 1.8 \times 10^{-10}$), (Supplementary Figure \ref{fig:chromosomeinteractions}). Relatedly, there was significant enrichment of \emph{trans}-SNPs being located within 250bp to kid/KIF22 binding sites ($p = 8.4 \times 10^{-28}$) which are involved in intracellular chromosome transport \cite{Miki2001}.

% \subsection{Contribution of epistatic variance}
Though we present many instances of epistasis, quantifying its relative importance to complex traits in humans remains an open question. In this study we are able to identify 238 gene expression traits with at least one significant interaction given our experiment-wide threshold. How does this compare to the number of traits influenced by additive effects? The BSGS dataset has been previously analysed for additive effects at all expression traits \cite{Powell2012}, and if we take all the additive eQTLs that were significant at the epistatic threshold of $p < 2.91 \times 10^{-16}$ we find that 453 gene expression levels out of the 7339 analysed had at least one significant expression quantitative trait locus (eQTL). Therefore it can be argued that the number of instances of detectable epistasis are substantial.

However in terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Ideally one would approach this question from a whole genome perspective \cite{Visscher2008} but this is intractable for non-additive variance components. Yet some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}. Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that in total they explain 1.73\% of the total phenotypic variance of all 7339 probes. By contrast, the epistatic variance from the interacting SNPs detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance. It must be noted that there are at least three caveats to this comparison. Firstly, the ratio of additive to epistatic variance may differ at different effect sizes. Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{Grey_graph_of_interactions_2_lists.pdf}
	\caption{\textbf{Discovery and replication of epistatic networks} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown, where black nodes represent SNPs and red nodes represent traits (gene expression probes). 316 Interactions had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data, but the remaining 118 interactions that did not replicate are depicted in grey. It is evident that a large proportion of the complex networks identified in the discovery set also exist in independent populations.}
	\label{fig:fireworks}
\end{figure}
\clearpage

\begin{figure}[H]
	\centering
	\includegraphics[width=2.5in]{gpbonfrep.pdf}
	\caption{\textbf{Replication of GP maps in two independent populations} The GP maps for each epistatic interaction that is significant at the Bonferroni level in both replication datasets are shown. Each GP map consists of nine tiles where each tile represents the expression level for that two-locus genotype class. Phenotypes are for gene transcript levels (dark coloured tiles = low expression, light coloured tiles = high expression). Columns of GP maps are for each independent population. Rows of GP maps are for each of 30 significantly replicated interactions, corresponding to the rows in Table \ref{tab:bonferroni}. There is a general trend of the GP maps replicating across all three datasets.}
	\label{fig:gpmaps}
\end{figure}
\clearpage

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{OverlapAnnotation}
	\caption{\textbf{Location of SNPs relative to genomic features} All SNPs within 5kb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic regions were covered by the 501 significant interactions. There is enrichment for \emph{cis}-acting SNPs in promotor regions, and \emph{cis}- and \emph{trans}-acting SNPs in intronic regions.}
	\label{fig:snplocations}
\end{figure}
\clearpage

\begin{figure}
	\centering
	\includegraphics[width=5in]{501_pairs_snp_list_Grid}
	\caption{\textbf{Tissue specific enrichment of SNPs in transcriptionally active regions} The locations of transcriptional activity can be predicted by chromatin marks, assayed by H3K4me3 \cite{Trynka2013}. Enrichment $p$-values are calculated using permutation analysis for 34 different cell types ($y$-axis) in four tissues (Rows of boxes). There is enrichment for \emph{cis}-acting SNPs in Haematopoeitic tissue types only. \emph{Trans}-acting SNPs have no tissue specificity.}
	\label{fig:h3k4me3}
\end{figure}
\clearpage
